---
title: "Calculating the Probability (a.k.a. *p-value*) of the Test-Statistic using R"
author: "Jeffrey M. Pisklak"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

- Critical Region ($\alpha = .05$) is shown in blue.  
- Solid vertical red line is the test-statistic.
- The probability of a test-statistic (i.e. p-value) is shown in red.
- Degrees of freedom are set to 4 in all examples.

<br>

## **Two-Tailed Tests**

### 1. If your **test-statistic is negative** (e.g., $T = -1.5$), then you are going to use . . .

```{r echo = TRUE}
pt(-1.5, df = 4) * 2
```

```{r echo = FALSE, fig.width = 4, fig.height = 3}
library(ggplot2)
x = seq(from = -4, to = 4, by = .001)
Density = dt(x, df = 4)
dist_df = data.frame(x, Density)

ggplot(dist_df, aes(x = x, y = Density)) +
  geom_line(size = 1) +
  
  #Title
  labs(title = "Two-Tailed",
        subtitle = "E.g., T = -1.5, df = 4") +
  
  #Left tail colour
  geom_ribbon(data = subset(dist_df, x < -1.5), aes(ymax = Density), 
                ymin = 0, fill = "red", colour = NA, alpha = .5) +

  #Right tail colour
  geom_ribbon(data = subset(dist_df, x > 1.5), aes(ymax = Density), 
                ymin = 0, fill = "red", colour = NA, alpha = .5) +
  
  geom_vline(xintercept = -1.5, colour = "red", size = 1) +
  
  annotate("text", x = -3, y= .25, label = "T = -1.5\np = 0.208") +
  
  #Axis Scale
  scale_x_continuous(name ="t-score", 
                    breaks = seq(-4, 4, 1) ) +

  #Left tail 95% Threshold
  geom_ribbon(data = subset(dist_df, x < qt(.05/2, df = 4)), 
              aes(ymax = Density),
              ymin = 0, fill = "blue", colour = NA) +

  #Right tail 95% Threshold
  geom_ribbon(data = subset(dist_df, x > qt(1-.05/2, df = 4)),
                            aes(ymax = Density),
              ymin = 0, fill = "blue", colour = NA)

```

<br>

### 2. If your **test-statistic is positive** (e.g., $T = 1.5$), then you are going to use . . .

```{r echo = TRUE}
pt(1.5, df = 4, lower.tail = FALSE) * 2
```

```{r echo = FALSE, fig.width = 4, fig.height = 3}
x = seq(from = -4, to = 4, by = .001)
Density = dt(x, df = 4)
dist_df = data.frame(x, Density)

ggplot(dist_df, aes(x = x, y = Density)) +
  geom_line(size = 1) +
  
  #Title
  labs(title = "Two-Tailed",
        subtitle = "E.g., T = +1.5, df = 4") +
  
  #Left tail colour
  geom_ribbon(data = subset(dist_df, x < -1.5), aes(ymax = Density), 
                ymin = 0, fill = "red", colour = NA, alpha = .5) +

  #Right tail colour
  geom_ribbon(data = subset(dist_df, x > 1.5), aes(ymax = Density), 
                ymin = 0, fill = "red", colour = NA, alpha = .5) +
  
  geom_vline(xintercept = 1.5, colour = "red", size = 1) +
  
  annotate("text", x = 3, y= .25, label= "T = +1.5\np = 0.208") +
  
  #Axis Scale
  scale_x_continuous(name ="t-score", 
                    breaks = seq(-4, 4, 1) ) +
  
  #Left tail 95% Threshold
  geom_ribbon(data = subset(dist_df, x < qt(.05/2, df = 4)),
              aes(ymax = Density),
              ymin = 0, fill = "blue", colour = NA) +

  #Right tail 95% Threshold
  geom_ribbon(data = subset(dist_df, x > qt(1-.05/2, df = 4)),
              aes(ymax = Density),
              ymin = 0, fill = "blue", colour = NA)

```

<br>

## **One-tailed tests**

### 3. If your **critical region is in the left tail** (i.e., $H_0 : \mu > x$) then you are going to use . . .

```{r echo = TRUE}
pt(1.5, df = 4)
```

```{r echo = FALSE, fig.width = 4, fig.height = 3}
x = seq(from = -4, to = 4, by = .001)
Density = dt(x, df = 4)
dist_df = data.frame(x, Density)

ggplot(dist_df, aes(x = x, y = Density)) +
  geom_line(size = 1) +
  
  #Title
  labs(title = "Two-Tailed",
        subtitle = "E.g., T = +1.5, df = 4") +
  
  #Left tail colour
  geom_ribbon(data = subset(dist_df, x < 1.5), aes(ymax = Density), 
                ymin = 0, fill = "red", colour = NA, alpha = .5) +

  geom_vline(xintercept = 1.5, colour = "red", size = 1) +
  
  annotate("text", x = 3, y = .25, label= "T = +1.5\np = 0.896") +
  
  #Axis Scale
  scale_x_continuous(name ="t-score", 
                    breaks = seq(-4, 4, 1) ) +
  
  #Left tail 95% Threshold
  geom_ribbon(data = subset(dist_df, x < qt(.05, df = 4)), 
              aes(ymax = Density),
              ymin = 0, fill = "blue", colour = NA)

```

<br>

### 4. If your **critical region is in the right tail** (i.e., $H_0 : \mu < x$) then you are going to use . . .

```{r echo = TRUE}
pt(1.5, df = 4, lower.tail = FALSE)
```

```{r echo = FALSE, fig.width = 4, fig.height = 3}
x = seq(from = -4, to = 4, by = .001)
Density = dt(x, df = 4)
dist_df = data.frame(x, Density)

ggplot(dist_df, aes(x = x, y = Density)) +
  geom_line(size = 1) +
  
  #Title
  labs(title = "Two-Tailed",
        subtitle = "E.g., T = +1.5, df = 4") +
  
  #Left tail colour
  geom_ribbon(data = subset(dist_df, x > 1.5), aes(ymax = Density), 
                ymin = 0, fill = "red", colour = NA, alpha = .5) +

  #Right tail colour
  #geom_ribbon(data = subset(dist_df, x > 1.5), aes(ymax = Density), 
   #             ymin = 0, fill = "blue", colour = NA, alpha = .5) +
  
  geom_vline(xintercept = 1.5, colour = "red", size = 1) +
  
  annotate("text", x = 3, y= .25, label= "T = +1.5\np = 0.104") +
  
  #Axis Scale
  scale_x_continuous(name ="t-score", 
                    breaks = seq(-4, 4, 1) ) +
  
  #Right tail 95% Threshold
  geom_ribbon(data = subset(dist_df, x > qt(.95, df = 4)), aes(ymax = Density), 
                ymin = 0, fill = "blue", colour = NA)

```